Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm: A Case Study

### 2022-05-03 Paper_Draft2/Draft2.tex
updating draft with all the parts now!

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{A variety of disciplines use risk assessment instruments to
help humans make data-driven decisions. Northpointe, a software company,
created an algorithmic risk assessment instrument known as the
Correctional Offender Management Profiling for Alternative Sanctions
(COMPAS). COMPAS uses various behavioral and psychological metrics
related to recidivism to assist justice systems in assessing a
defendant's potential recidivism risk. \citet{angwin2016machine}
published a ProPublica article in which they conclude that the racial
biases in the criminal justice system are reflected in the COMPAS
recidivism risk scores. In response, \citet{equivant_response_2016}
published a rebuttal on behalf of Northpointe defending the COMPAS
algorithm and refuting \citet{angwin2016machine}'s allegation of racial
bias. Using a human rights framework adopted from the organizations
Women at the Table and AI Fairness 360, we use debiasing algorithms and
fairness metrics to analyze the argument between Northpointe and
ProPublica and determine whether and to what extent there is racial bias
in the COMPAS algorithm. All four group fairness metrics determine that
the COMPAS algorithm favors white defendants over Black defendants.}

@@

Due to the proprietary nature of the COMPAS algorithm, it is unknown how
exactly these recidivism risk scores are calculated. However, a
\href{https://www.documentcloud.org/documents/2702103-Sample-Risk-Assessment-COMPAS-CORE\#document/p5/a296598}{sample
COMPAS Risk Assessment Survey} has been made publicly available,
revealing the algorithm's input information. \citet{angwin2016machine}
critiques this survey for using proxy variables for race that do not
explicitly factor in a defendant's race but heavily imply it, allowing
Northpointe to claim that their algorithm is free of racial bias. For
example, the COMPAS risk assessment survey asks screeners to speculate
if a defendant might be affiliated with a gang. It also asks if a
defendant has any friends or family members who have been crime victims.
Although these questions do not directly ask about race, they do not
take into account the pervasive nature of systemic racism that
infiltrates every aspect of the lives of marginalized people, thereby
indirectly asking about race.

\citet{angwin2016machine} analyzes the methods and algorithms used by
Northpointe in their COMPAS risk score assessment algorithm and uncovers
racial biases in defendants' scores \citep{angwin2016machine}. They find
that ``the algorithm {[}is{]} somewhat more accurate than a coin flip,''
a worrisome level of accuracy given the potential impact its
determinations may have on real people's lives.
\citet{angwin2016machine} specifically investigate the distribution of
COMPAS scores by decile among Black and white defendants. They write:
``The analysis also {[}shows{]} that even when controlling for prior
crimes, future recidivism, age, and gender, Black defendants {[}are{]}
45 percent more likely to be assigned higher risk scores than white
defendants'' \citep{larson2016we}. After examining the fairness metric
statistical parity difference, \citet{angwin2016machine} conclude that
the algorithm is racially biased \citep{larson2016we}.

@@

\href{https://www.womenatthetable.net/}{Women at the Table}, the sponsor
organization for this project, is ``a growing, global gender equality \&
democracy CSO based in Geneva, Switzerland focused on advancing feminist
systems change by using the prism of technology, innovation \& AI
exercising leverage points in technology, the economy, sustainability \&
democratic governance.'' We are collaborating with the organization on
its AI \& Equality \citep{noauthor_ai_nodate} initiative, tasked with
de-biasing the COMPAS algorithm \citep{aif360-oct-2018} and producing a
corresponding data story that will be added to its library.

Our project builds on Women at the Table's various de-biasing algorithms
used in its AI \& Equality Human Rights Toolbox to conduct our own
analyses on the COMPAS data set. Based on this analysis, we employ a
human rights framework to contribute to the ProPublica and Northpointe
debate and investigate whether or to what extent there is racial bias in
the COMPAS algorithm. With a solid understanding of the two sides, we
aim to pinpoint the shortcomings of both arguments and correct them in
our analyses. We will use various de-biasing techniques and fairness
metrics to evaluate the level of bias present in the COMPAS data and our
algorithm. We will summarize our results using the JupyterNotebook
framework from Women at the Table, to be used by members of the
organization to teach in a workshop setting. We hope that our findings
will highlight the importance of checking statistical analyses using
varied methods and contribute to the ongoing discussion of the effects
of machine biases in the justice system.

@@

The data we are using for this project is the COMPAS General Recidivism
Risk Scores data set from the AI Fairness 360 (AIF360) toolkit
\citep{aif360-oct-2018}, which does the same initial pre-processing as
ProPublica. The raw data has 6,167 rows and each row represents an
arrest charge for a defendant. AIF360's COMPAS data includes the
defendant's age, race, sex, prior charges, what they were charged with,
and whether or not the defendant ultimately recidivated within a
two-year period after their arrest. For the purposes of our project,
which endeavors to evaluate the differing effects of the COMPAS
algorithm on white defendants versus Black defendants, we have filtered
the data to only include individuals whose race is listed as Caucasian
or African-American. Our data therefore has 5,273 rows (Figure
\ref{fig:table snip}), with the distributions of age (Figure
\ref{fig:age plot}), prior charges (Figure \ref{fig:prior plot}), and
two-year recidivism by race (Figure \ref{fig:recidivism mosaic}) and
(Table \ref{tab:recid table}) shown below.

@@

The AIF360 toolkit contains various group and individual fairness
metrics as well as pre-processing, in-processing, and post-processing
algorithms that we used to debias the COMPAS algorithm
\citep{aif360-oct-2018}. We researched the definitions and applications
of different fairness metrics \citep{ashokan2021fairness} to determine
which metric would be most appropriate for our project. We chose to look
at four group fairness metrics instead of individual fairness metrics.
Group fairness metrics take into account the attributes of a whole group
as opposed to just one individual in the group, allowing us to represent
systemic issues. In general, group fairness metrics require that the
unprivileged group is treated similarly to the privileged group, whereas
individual fairness metrics require individuals to be treated
consistently \citep{kypraiou_what_2021}. Group and individual metrics
work in opposition of one another, meaning that when group fairness
improves, individual fairness gets worse \citep{kypraiou_what_2021}.

@@

For the next step of our experiment, we need to determine where in the
data science pipeline we can mitigate the most bias, using
pre-processing, in-processing, and post-processing de-biasing
algorithms. These are all based on using predictive models to figure out
how we can ``fix'' the bias that is present.

@@

In-processing mitigates bias in classifiers while building a model. A
classifier ``is an algorithm that automatically orders or categorizes
data into one or more sets'' \citep{baxter2021AI}. The in-processing
technique we use is the prejudice remover algorithm, which accounts for
the fairness metric as part of the input and returns a classifier
optimized by that particular metric. In order to do this, we first
needed to convert our data frame into a data type called a
BinaryLabelDataset. Similarly to pre-processing, we compared the results
of our in-processing methods with both the baseline and the
pre-processing to gauge which method so far has better individual or
group fairness.

The prejudice remover is a method for reducing indirect prejudice
(i.e.~how COMPAS is racially biased because it uses proxy variables for
race). The prejudice remover implements two different regularizers, one
to avoid overfitting and one to enforce fair classification. The
prejudice remover regularizer works by minimizing the prejudice index, a
mathematical equation for quantifying fairness defined by Kamishima et.
al.~This in turn enforces a classifier's independence from sensitive
information (e.g., race).

Our last approach, post-processing bias mitigation, is used after
training a model. Post-processing algorithms equalize the outcomes
(i.e., predicted classification values) to mitigate bias instead of
adjusting the classifier or the training data \citep{baxter2021AI}. We
use calibrated equalized odds, which ``optimizes over calibrated
classifier score outputs to find probabilities with which to change
output labels with an equalized odds objective'' {[}insert aif360
website citation here{]}. An equalized odds objective constrains
classification algorithms such that no error type (false-positive or
false-negative) disproportionately affects any population subgroup; both
groups, in our case both white and Black defendants, should have the
same false-positive and false-negative rates. Through the calibrated
equalized odds method, we want to decrease bias while also maintaining
calibration \citep{pleiss2017fairness}.
\href{https://medium.com/analytics-vidhya/calibration-in-machine-learning-e7972ac93555}{Calibration}
refers to improving a model so that the distribution of predicted
outcomes is similar to the distribution of observed probability in the
training data.

@@

Equal opportunity difference is -0.41. The difference of true positive
rates between the Black and white groups is -0.41. A value less than 0
indicates a benefit to the privileged group, so the algorithm unfairly
benefits white defendants. The value is substantially less than -0.1,
which indicates that the algorithm benefits white defendants.

All four group fairness metrics determine that the COMPAS algorithm
favors white defendants over Black defendants. Although the magnitudes
of the various fairness metrics are different, all of the fairness
metrics are not within their respective fairness thresholds. Our goal is
to use pre-processing, in-processing, and post-processing algorithms in
the AIF360 toolkit to see if we can make COMPAS fair at all.

\begin{longtable}[]{@{}llll@{}}
\caption{Results of Baseline Fairness Metric Analysis
\label{tab:baseline metrics table}}\tabularnewline
\toprule
Fairness Metric & Ideal Value & Baseline Value & Benefited
Group\tabularnewline
\midrule
\endfirsthead
\toprule
Fairness Metric & Ideal Value & Baseline Value & Benefited
Group\tabularnewline
\midrule
\endhead
Statistical Parity Difference & 0 & -0.14 & White
Defendants\tabularnewline
Disparate Impact Ratio & 1 & 0.47 & White Defendants\tabularnewline
Average Odds Difference & 0 & -0.44 & White Defendants\tabularnewline
Equal Opportunity Difference & 0 & -0.41 & White
Defendants\tabularnewline
\bottomrule
\end{longtable}

\hypertarget{pre-processing-approach}{%
\subsection{Pre-Processing Approach}\label{pre-processing-approach}}

After running the reweighing algorithm, our fairness metrics are -0.015
for statistical parity difference, 0.015 for equal opportunity
difference, 0.014 for average odds difference, and 0.98 for disparate
impact ratio. Overall, these fairness metrics show that the
pre-processing algorithm reweighing improves the bias in the COMPAS
algorithm.

>>>>

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{A variety of disciplines use risk assessment instruments to
help humans make data-driven decisions. Northpointe, a software company,
created an algorithmic risk assessment instrument known as the
Correctional Offender Management Profiling for Alternative Sanctions
(COMPAS). COMPAS uses various behavioral and psychological metrics
related to recidivism to assist justice systems in assessing a
defendant's potential recidivism risk. Angwin et al.~published a
ProPublica article in which they conclude that the racial biases in the
criminal justice system are reflected in the COMPAS recidivism risk
scores. In response, Dieterich et al.~published a rebuttal on behalf of
Northpointe defending the COMPAS algorithm and refuting Angwin et al.'s
allegation of racial bias. Using a human rights framework adopted from
the organizations Women at the Table and AI Fairness 360, we use
debiasing algorithms and fairness metrics to analyze the argument
between Northpointe and ProPublica and determine whether and to what
extent there is racial bias in the COMPAS algorithm. All four group
fairness metrics determine that the COMPAS algorithm favors white
defendants over Black defendants. Our research found that the pre and
post-processing bias mitigation algorithms, specifically reweighing and
calibrated equalized odds, are the most effective at improving
fairness.}

@@

Due to the proprietary nature of the COMPAS algorithm, it is unknown how
exactly these recidivism risk scores are calculated. However, a
\href{https://www.documentcloud.org/documents/2702103-Sample-Risk-Assessment-COMPAS-CORE\#document/p5/a296598}{sample
COMPAS Risk Assessment Survey} has been made publicly available,
revealing the algorithm's input information. \citet{angwin2016machine}
critiques this survey for using proxy variables for race that do not
explicitly factor in a defendant's race but heavily imply it, allowing
Northpointe to claim that their algorithm is free of racial bias. For
example, the COMPAS risk assessment survey asks screeners to speculate
if a defendant might be affiliated with a gang. It also asks if a
defendant has any friends or family members who have been crime victims
\citep{Angwin2016Sample}. Although these questions do not directly ask
about race, they do not take into account the pervasive nature of
systemic racism that infiltrates every aspect of the lives of
marginalized people, thereby indirectly asking about race.

\citet{angwin2016machine} analyzes the methods and algorithms used by
Northpointe in their COMPAS risk score assessment algorithm and uncovers
racial biases in defendants' scores \citep{angwin2016machine}. They find
that ``the algorithm {[}is{]} somewhat more accurate than a coin flip,''
a worrisome level of accuracy given the potential impact its
determinations may have on real people's lives. Angwin et
al.~specifically investigate the distribution of COMPAS scores by decile
among Black and white defendants. They write: ``The analysis also
{[}shows{]} that even when controlling for prior crimes, future
recidivism, age, and gender, black defendants {[}are{]} 45 percent more
likely to be assigned higher risk scores than white defendants''
\citep{larson2016we}. After examining the fairness metric statistical
parity difference, Angwin et al.~conclude that the algorithm is racially
biased \citep{larson2016we}.

@@

\href{https://www.womenatthetable.net/}{Women at the Table}, the sponsor
organization for this project, is ``a growing, global gender equality \&
democracy CSO based in Geneva, Switzerland focused on advancing feminist
systems change by using the prism of technology, innovation \& AI
exercising leverage points in technology, the economy, sustainability \&
democratic governance.'' We are collaborating with the organization on
its AI \& Equality \citep{noauthor_ai_nodate} initiative, tasked with
debiasing the COMPAS algorithm \citep{aif360-oct-2018} and producing a
corresponding data story that will be added to its library.

Our project builds on Women at the Table's various debiasing algorithms
used in its AI \& Equality Human Rights Toolbox to conduct our own
analyses on the COMPAS data set. Based on this analysis, we employ a
human rights framework to contribute to the ProPublica and Northpointe
debate and investigate whether or to what extent there is racial bias in
the COMPAS algorithm. With a solid understanding of the two sides, we
aim to pinpoint the shortcomings of both arguments and correct them in
our analyses. We will use various debiasing techniques and fairness
metrics to evaluate the level of bias present in the COMPAS data and our
algorithm. We will summarize our results using the JupyterNotebook
framework from Women at the Table, to be used by members of the
organization to teach in a workshop setting. We hope that our findings
will highlight the importance of checking statistical analyses using
varied methods and contribute to the ongoing discussion of the effects
of machine biases in the justice system.

@@

The data we are using for this library addition is the COMPAS General
Recidivism Risk Scores data set from the
\href{https://aif360.readthedocs.io/en/stable/modules/generated/aif360.datasets.CompasDataset.html\#aif360.datasets.CompasDataset}{AI
Fairness 360} (AIF360) toolkit. The AIF360 toolkit builds on the dataset
released by ProPublica collaborators Larson et al.~created to examine
the racial bias and the true outcomes of the recidivism risk scores in
the COMPAS algorithm for the initial
\href{https://www.propublica.org/article/machine-bias-risk-assessments-in-criminal-sentencing}{``Machine
Bias''} article. For this, Larson et al.~obtained two years worth of
COMPAS scores from the Broward County Sheriff's Office in Florida, as
well as the corresponding intake information for each defendant
including but not limited to name, sex, race, age, and charge degree and
description. They also obtained data about whether defendants actually
recidivated or not in the two year period following their initial COMPAS
score assessment. AIF360 then processed the data with the same
procedures that Larson et al.~followed for their analysis.

The raw data that we use from the AIF360 has 6,167 rows, where each row
represents an arrest charge for a defendant. AIF360's COMPAS data
includes the defendant's age, race, sex, what they were charged with,
and whether or not the defendant ultimately recidivated within a
two-year period after their arrest.

This exploration aims to evaluate anti-Black algorithmic bias and the
differing effects of the COMPAS algorithm between white and Black
defendants; as such, we filter the data to only include individuals
whose race is listed as Caucasian or African-American. Our data
therefore has 5,723 rows with information on defendant race, age,
gender, prior crimes, and two-year recidivism rate. The average
African-American defendant in our dataset is 29 years old, male, and has
committed two prior crimes. The average Caucasian defendant in our
dataset is 35 years old, male, and has committed one prior crime. The
median defendant for both races do not have any juvenile convictions.

@@

The AIF360 toolkit contains various group and individual fairness
metrics as well as pre-processing, in-processing, and post-processing
algorithms that we used to debias the COMPAS algorithm
\citep{aif360-oct-2018}. Fairness metrics are mathematical measures of
whether an algorithm treats members of different groups (such as racial
or gender groups) equally. An algorithm has no understanding of the
historical oppression of certain groups and how such bias is baked into
the data on which the algorithm is trained. Thus, fairness metrics
provide a method of evaluating an algorithm's level of bias towards or
against the unprivileged group versus the privileged group. We
researched the definitions and applications of different fairness
metrics \citep{ashokan2021fairness} to determine which metric would be
most appropriate for our project.

Fairness is subjective; one person might consider an algorithm fair if
groups are given the same treatment, while someone else might only
consider the algorithm fair if groups as a whole receive the same
outcomes. In our research we choose to use the definition of group
fairness as our definition for assessing fairness in the processing
approach. Group fairness metrics take into account the attributes of a
whole group as opposed to just one individual in the group, allowing us
to represent systemic issues. In general, group fairness metrics require
that the unprivileged group is treated similarly to the privileged
group, whereas individual fairness metrics require individuals to be
treated consistently \citep{kypraiou_what_2021}. Group and individual
metrics work in opposition of one another, meaning that when group
fairness improves, individual fairness gets worse
\citep{kypraiou_what_2021}. We chose the following four group fairness
metrics to evaluate our models.

@@

For the next step of our experiment, we sought to determine where in the
data science pipeline we can mitigate the most bias, using
pre-processing, in-processing, and post-processing debiasing algorithms.
These are all based on using predictive models to figure out how we can
``fix'' the bias that is present.

\hypertarget{pre-processing}{%
\subsection{Pre-Processing}\label{pre-processing}}

@@

\hypertarget{in-processing}{%
\subsection{In-Processing}\label{in-processing}}

In-processing mitigates bias in classifiers while building a model. A
classifier ``is an algorithm that automatically orders or categorizes
data into one or more sets'' \citep{baxter2021AI}. The in-processing
technique we use is the prejudice remover algorithm, which accounts for
the fairness metric as part of the input and returns a classifier
optimized by that particular metric. In order to do this, we first
needed to convert our data frame into a data type called a
BinaryLabelDataset.

The prejudice remover is a method for reducing indirect prejudice (i.e.,
how COMPAS is racially biased because it uses proxy variables for race).
The prejudice remover implements two different regularizers, one to
avoid overfitting and one to enforce fair classification
\citep{kamishima2012fairness}. The prejudice remover regularizer works
by minimizing the prejudice index, a mathematical equation for
quantifying fairness defined by \citet{kamishima2012fairness}. This in
turn enforces a classifier's independence from sensitive information
(e.g., race). Similarly to pre-processing, we compared the results of
our in-processing methods with both the baseline and the pre-processing
model to gauge which method so far has the better group fairness.

\hypertarget{post-processing}{%
\subsection{Post-Processing}\label{post-processing}}

Our last approach, post-processing bias mitigation, is implemented after
training a model. Post-processing algorithms equalize the outcomes
(i.e., predicted recidivism values) to mitigate bias instead of
adjusting the classifier or the training data \citep{baxter2021AI}. We
use calibrated equalized odds, which ``optimizes over calibrated
classifier score outputs to find probabilities with which to change
output labels with an equalized odds objective''
\citep{aif360-oct-2018}. An equalized odds objective constrains
classification algorithms such that no error type (false-positive or
false-negative) disproportionately affects any population subgroup; both
groups, in our case both white and Black defendants, should have the
same false-positive and false-negative rates \citep{pleiss2017fairness}.
Through the calibrated equalized odds method, we want to decrease bias
while also maintaining calibration \citep{pleiss2017fairness}.
\href{https://medium.com/analytics-vidhya/calibration-in-machine-learning-e7972ac93555}{Calibration}
refers to improving a model so that the distribution of predicted
outcomes is similar to the distribution of observed probability in the
training data.

@@

Equal opportunity difference is -0.41. The difference of true positive
rates between the Black and white groups is -0.41. A value less than 0
indicates a benefit to the privileged group, so the algorithm benefits
white defendants. The value is substantially less than -0.1, which
indicates that the algorithm is unfair.

All four group fairness metrics determine that the COMPAS algorithm
favors white defendants over Black defendants. Although the magnitudes
of the various fairness metrics are different, none of the metrics are
within their respective fairness thresholds. Our goal is to use
pre-processing, in-processing, and post-processing algorithms in the
AIF360 toolkit to see if we can make COMPAS fair at all.

Our confusion matrix for the baseline model indicates that the false
positive and false negative percentages are 16.68\% and 22.06\%, which
are the lowest values for this matrix. The highest value is the value of
true positives, at 35.41\%, showing that of the total number of
predictions, 35.41\% were correct predictions of recidivism.

\begin{longtable}[]{@{}llll@{}}
\caption{Results of Baseline Fairness Metric Analysis
\label{tab:baseline metrics table}}\tabularnewline
\toprule
Fairness Metric & Ideal Value & Baseline Value & Benefited Group \\
\midrule
\endfirsthead
\toprule
Fairness Metric & Ideal Value & Baseline Value & Benefited Group \\
\midrule
\endhead
Statistical Parity Difference & 0 & -0.14 & White Defendants \\
Disparate Impact Ratio & 1 & 0.47 & White Defendants \\
Average Odds Difference & 0 & -0.44 & White Defendants \\
Equal Opportunity Difference & 0 & -0.41 & White Defendants \\
\bottomrule
\end{longtable}

\hypertarget{pre-processing-approach-reweighing}{%
\subsection{Pre-Processing Approach:
Reweighing}\label{pre-processing-approach-reweighing}}

After running the reweighing algorithm, our fairness metrics are -0.015
for statistical parity difference, 0.015 for equal opportunity
difference, 0.014 for average odds difference, and 0.98 for disparate
impact ratio. All of these values now fall within the margins of
fairness (see reweighing metrics plot). These values make sense because
the reweighing algorithm chooses different weights depending on whether
an attribute is protected. This does an effective job of eliminating the
bias, shown by each of the values in the graph being within the ``fair''
range of the particular fairness metric. Overall, these fairness metrics
show that the reweighing algorithm improves the bias in the COMPAS
algorithm.

The highest percentage in this confusion matrix is for the percentage of
true negatives, defendants who were predicted to not recidivate and
actually did not, at 36.24\%, 0.83\% higher than the true negatives in
the baseline model. This means that the reweighing model marginally
improved the baseline model accuracy in predicting people who did not
recidivate. However, the percentage of true positives, or those who were
predicted to recidivate and did recidivate, is 24.72\%, 1.13\% lower
than that of the baseline model. This means that the reweighing model
slightly lowered the accuracy of the baseline model in predicting people
who did recidivate. As a result, the overall reweighing model accuracy
remains about the same as the baseline model accuracy. The false
negative and false positive percentages, or the predictions that proved
to be incorrect, are 23.20\% and 15.85\%, respectively. The re-weighing
model increased the number of false negatives by 1.14\% and decreased
the number of false positives by 0.83\%. We are most concerned about the
false positives, which indicate defendants who are predicted to
recidivate and do not actually recidivate. While there is a slight
improvement in the percent of false positives, it is still relatively
high at 15.85\%, indicating that we should further precede with
additional bias mitigation techniques.

\hypertarget{in-processing-approach-prejudice-remover}{%
\subsection{In-Processing Approach: Prejudice
Remover}\label{in-processing-approach-prejudice-remover}}

Our fairness metrics after running the prejudice remover are: 0.367 for
statistical parity difference, 0.321 for equal opportunity difference,
0.344 for average odds difference, and 3.3 for disparate impact ratio.
Like the model performance metrics suggested, the prejudice remover
approach resulted in an increased benefit for Black defendants. As the
orange bars on these plots show, the values of the fairness metrics have
reversed from their baseline values. Now all four metrics suggest an
unfair advantage for Black defendants. Thus, this approach removed the
model's prejudice against Black people, but it did not result in a
``fair'' model.

The highest percentage in this confusion matrix is for false positives,
defendants who the model predicted to recidivate and actually did not,
at 37.76\%. This false positive percentage concerns us because we do not
want defendants who do not recidivate to have unfairly long sentences
due to their (incorrectly) predicted recidivism. The number of true
positives, defendants who the model predicted to recidivate and actually
recidivated, is lower at 23.88\%. The percentage of false negatives is
lower, whereas the true negative is low at 14.33\%. This indicates that
the model does not do a good job in predicting the defendants who don't
recidivate. It is very likely that it will incorrectly predict that
someone will recidivate.

After the prejudice remover, the accuracy of the model is fairly similar
across the board of races, but the model's accuracy is only 38.21\%,
which is much lower than the baseline and reweighing models. This
accuracy score means that the model makes accurate predictions only 38\%
of the time.

\hypertarget{post-processing-approach-calibrated-equalized-odds}{%
\subsection{Post-Processing Approach: Calibrated Equalized
Odds}\label{post-processing-approach-calibrated-equalized-odds}}

After running the calibrated equalized odds algorithm, our fairness
metrics are: 0.134 for statistical parity difference, 0.088 for equal
opportunity difference, 0.146 for average odds difference, and 1.155 for
disparate impact ratio. As we can see from the orange bars, the
calibrated equalized odds approach results in all four metrics now
suggesting a benefit for the originally unprivileged group, Black
defendants. Though the values for statistical parity difference and
average odds difference are slightly above the range of what is
considered fair, the margin is much smaller than the original margin
between the value and the range of fairness (as indicated by the blue
bars). Thus, the calibrated equalized odds approach successfully
counteracts the bias against Black defendants and results in a mostly
fair model.

The highest percentage in this confusion matrix is for true negatives,
defendants who the model predicted to not recidivate and actually did
not, at 50.04\%. The number of true positives, defendants who the model
predicted to recidivate and actually recidivated, is much lower at
3.11\%. The percentage of false negatives is somewhat high for this
baseline model, whereas the amount of false positives is very low at
2.05\%. This indicates that the model does a good job in predicting the
defendants who don't recidivate. It is very unlikely that it will
incorrectly predict that someone will recidivate. Though the accuracy
score for the model is 0.53, lower than the baseline, this model is much
better at identifying true negatives.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

\hypertarget{limitations}{%
\subsection{Limitations}\label{limitations}}

Our project tries to mitigate bias within existing data, rather than
within the methods used to collect this data. The data are collected
using a risk assessment survey, where defendants are asked a series of
questions (see introduction) which are supposedly used to determine
whether someone will recidivate or not. Many of these questions involve
proxy variables for race. The creation and facilitation of this survey
can therefore be assessed for bias mitigation as the methods used may
potentially privilege some groups over others. As a result, the bias
exists even before values are collected, limiting the debiasing work
able to be done on the output values.

Furthermore, AIF360 processed the data we used in this project the same
way ProPublica processed their data. AIF360 simplifies the original
defendant data and their COMPAS scores to make the data easier to
analyze. However, this initial processing may lose some details captured
in the original raw data. As we have noted throughout this paper, bias
can be introduced in any step of the data science pipeline and AIF360's
initial data processing step is no exception.

AIF360 contains an array of processing methods that have been written by
various scholars of fairness \citep{aif360-oct-2018}. These methods are
each distinct and apply a different technique to mitigate bias. Our
results only contain three processing methods, one for each step in
processing. This choice was based on time constraints as well as our
working environment. For example, the code for most of the other
in-processing methods would not run for us. In-processing on its own
works to debias an algorithm, not the input or output values, so some of
those methods may not work on certain algorithms. We had similar
challenges with pre and post-processing, where the methods were
challenging to run and would require more time to be spent debugging in
order to make them work. Had we used alternative processing methods, our
results would be quite different, as the method changes apply a
different mathematical technique to the input.

\hypertarget{future-work}{%
\subsection{Future Work}\label{future-work}}

In this project, we have only focused on a few select fairness metrics
and a few debiasing algorithms. We looked at group fairness metrics
because we specifically wanted to mitigate systemic racial bias in the
COMPAS algorithm, but other researchers could compare how well debiasing
algorithms work using individual fairness metrics. Future work could
also include conducting more tests using other debiasing algorithms to
see if the COMPAS algorithm could be made more fair. One significant
limitation of our work is that some of the in-processing algorithms
provided in the AIF360 toolkit were difficult to integrate into our code
and we were unable to run many of them. The processing methods we chose,
though effective, were based on our work environments, and there are
other existing algorithms that attempt to mitigate bias.

We only looked at defendants labeled as Caucasian and African-American
in our dataset, which diminishes the generalizability and nuance of our
results. Comparing how COMPAS assesses defendants of other races would
be an incredibly relevant extension of our existing work. Furthermore,
\citet{larson2016we} also found that age was the most predictive factor
of a higher risk score: ``Defendants younger than 25 years old were 2.5
times as likely to get a higher score than middle aged offenders, even
when controlling for prior crimes, future criminality, race and gender''
\citep{larson2016we}. In addition, female defendants were more likely to
receive a higher recidivism risk score than male defendants when
controlling for the aforementioned factors \citep{larson2016we}. Thus,
some extensions of our work could include attempting to remove age and
gender bias and determining the most effective processing algorithms to
do so.

The dataset we are working with only contains defendants' criminal
history records from Broward County, Florida. Many other states,
including New York and Massachusetts, use COMPAS to calculate recidivism
scores, and a further study could apply this paper's methods to the data
coming out of other states. By conducting experiments on multiple states
the baseline model would be trained on a larger and more diverse set of
data, which could improve its accuracy. In addition, by adding more
geographically diverse data to our set we can start to create more
generalizable conclusions about our results. However, future researchers
must keep in mind the real-life consequences of recidivism risk
algorithms like COMPAS and how research on methods of debiasing
algorithms could potentially be utilized to justify the continued use of
biased algorithms.

\hypertarget{final-thoughts}{%
\subsection{Final Thoughts}\label{final-thoughts}}

Through our methods, we concluded that the pre-processing and
post-processing methods mitigate bias and increase fairness most
successfully. The in-processing technique was effective at mitigating
bias, but did not result in a mathematically fair model. However, these
techniques only treat the symptoms of racial bias in the justice system,
as opposed to addressing the root cause, systemic racism. We need to
question why these flawed algorithms exist and why they can be used to
determine someone's fate in the justice system. While we may have
methods for reducing bias algorithms such as COMPAS, the focus should be
on addressing the over-policing and over-criminalization of Black
communities that result in biased data and algorithms.

\hypertarget{ethical-statement}{%
\subsection{Ethical Statement}\label{ethical-statement}}

Throughout this project, we endeavored to debias an incredibly powerful
algorithm that can change the course of someone's life. However, we want
to recognize that debiasing methods are not the most effective or
beneficial way to uproot bias within the American justice system. The
problems within the justice system are much more complex than an
algorithm and are rooted in the United States's history of racism. In
addition, algorithmic bias is much more complex than the algorithm
itself; algorithmic bias comes from the people who produce the
algorithm. We all have our own beliefs, biases, and situated knowledge
that impact and subsequently limit everything we create.

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

We would like to thank our sponsor organization Women At The Table and
Professor Ben Baumer for introducing us to this incredibly meaningful
work, supporting us, and answering the many questions we had throughout
this semester-long project. We would specifically like to thank Sofia
Kypraiou for her mentorship and patience throughout this process.
